Phenotypic evolution is more constrained in simpler food webs

### 2019-01-22 manuscript.Rmd
rerun reproducible analyses for manuscript
minor edits to manuscript.

```{r setup ms, include=FALSE}
## Reproduce analyses ----
rmarkdown::render("analyses/reproduce_analyses.Rmd")

@@

We conducted our study within a four-year old common garden experiment of coastal willow (*Salix hookeriana*) located at Humboldt Bay National Wildlife Refuge (HBNWR) (40 40'53"N, 124 12'4"W) near Loleta, California, USA. This common garden consists of 26 different willow genotypes that were collected from a single population of willows growing around Humboldt Bay. Stem cuttings of each genotype (25 replicates per genotypes) were planted in a completely randomized design in two hectares of a former cattle pasture at HBNWR. Willows at our study site begin flowering in February and reach their peak growth in early August. During this study, willows had reached 5 - 9m in height. Further details on the genotyping and planting of the common garden are available in @Barbour2015.

@@

We collected data on three different traits that we expected to influence larval survival based on our previous work [@Barbour2016] and other work with gall midges [@Weis1983; @Heath2018]. First, we measured gall diameter as the size of each gall chamber to the nearest 0.01 mm at its maximum diameter (perpendicular to the direction of plant tissue growth). Our previous work has shown that larger galls are associated with higher survival [@Barbour2016]. Second, we measured clutch size by counting the number of chambers in each gall [@Weis1983; @Heath2018]. All larvae collected from the same multi-chambered gall were scored with the same clutch size. Third, we measured oviposition preference as the density of larvae observed on a plant in an independent survey. We did this by randomly sampling five branches per tree and counting the number of individual gall chambers (number of larvae). We then converted these counts to a measure of larval density per 100 shoots by counting the number of shoots on the last branch we sampled. All larvae collected from the same plant were scored with the same oviposition preference. Measuring larval densities on plants in the field is a common method for measuring oviposition preference [@Gripenberg2010]; however, caution must be taken in inferring 'preference' as larval densities can be influenced by processes other than preference [@Singer1986]. Fortunately, a couple of features of our study system suggest that larval densities may be a good proxy for oviposition preference. For example, since our data comes from a randomized placement of plant genotypes in a common garden, there is no consistent bias in which plant genotypes females are exposed to while searching for oviposition sites. Also, egg predation is a minor source of mortality for galling insects in general [@Hawkins1997]; therefore, we do not expect any prior egg predation to bias our estimates of observed larval densities.

@@

Our analyses consisted of three parts. First, we used generalized linear mixed models (GLMM) to quantify selection surfaces \textemdash linear and nonlinear relationships between absolute fitness ($W$) and phenotypic traits ($z_i$) of individuals \textemdash in each food-web treatment. Second, we translated selection surfaces into the scale of relative fitness ($w$) in order to calculate selection gradients. Third, we used our estimates of selection gradients to characterize the slope and curvature of the adaptive landscape. Note that inferring adaptive landscapes from selection surfaces assumes that trait distributions are multivariate normal [@Lande1983]. To approximate this assumption, we log-transformed clutch size and square-root transformed oviposition preference. We then scaled all phenotypic traits (mean=0 and SD=1) in order to calculate standardized selection gradients that were comparable across traits and with other studies of natural selection.

@@

All analyses and visualizations were conducted in R [@R2018]. Unless otherwise noted, we report mean estimates of trait-fitness relationships and selection gradients with 95% confidence intervals in brackets. Note that for visualizing the adaptive landscape we restrict trait axes to $\pm 1$ SD of the mean trait value. This emphasizes the fact that we can only reliably estimate the shape of the adaptive landscape near the mean phenotype of the population [@Arnold2001]. We also plot mean larval survival on a natural log scale to accurately reflect the mathematical definition of the adaptive landscape [@Arnold2003].

@@

\indent We found that more gall midge traits experienced directional selection in the simple (3 of 3) vs. complex food-web treatment (1 of 3)(table \ref{Table:Gradients}). For example, we observed directional selection for smaller clutch sizes in the simple food web, but there was no evidence of selection acting on this trait in the complex food web (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This absence of selection appeared to be a result of conflicting selection pressures imposed by each guild of parasitoids (fig. \ref{fig:UV_Landscape}). Specifically, when we subset our data to focus on differences between parasitoid guilds, we found that larval parasitoids actually impose directional selection for larger clutch sizes ($\beta_{\text{Clutch}}$=
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$mean,2)` 
[`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.high,2)`]). In the simple food web, we also observed clear evidence of directional selection for midges to avoid ovipositing on plants with high densities of conspecifics (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}B); however, this relationship was weaker in complex food webs (table \ref{Table:Gradients}). This was likely a result of larval parasitoids imposing greater mortality on egg parasitoids at high gall midge densities (see Selection on the extended phenotype of egg parasitoids section), and thus a less than additive effect on gall midges. Chamber diameter experienced positive directional selection in both food-web treatments, but selection was `r round(((filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean - filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) / filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) * 100, 0)`% higher in complex food webs (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}). This was not due to any difference between egg and larval parasitoids (fig. \ref{fig:UV_Landscape}), but was due to the lower average survival in complex food webs (contrast `r round(filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$mean*-1, 2)`
[`r round(filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.low*-1, 2)`,
`r round(filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.high*-1, 2)`]).

@@

\begin{table}[h]
\caption{Standardized selection gradients acting on gall midges in complex vs. simple food webs.}
\label{Table:Gradients}
\centering
\begin{tabular}{lccc}
\\ 
\hline
\textbf{Selection gradient} & \textbf{Complex} & \textbf{Simple} & \textbf{Contrast}  \\ 
\hline
$\beta_{\text{Diam}}$ & 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.high,2)`] }& 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.high,2)`] }&

`r round(filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.high,2)`] \\

@@

```{r MV_Landscape, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:MV_Landscape}Two dimensional view of adaptive landscapes of gall midge phenotypes in complex vs. simple food webs. Each panel corresponds to a different combination of phenotypic traits: clutch size and chamber diameter (A); clutch size and oviposition preference (B); oviposition preference and chamber diameter (C). Arrows represent mean estimates of directional selection gradients, while contours represent predicted larval survival of the mean phenotype (i.e., all traits = 0) in each food-web treatment. Notice that arrows point more toward a corner of the adaptive landscape for each combination of traits in the simple vs. complex food-web treatment. This indicates that trait integration (covariance) is more strongly favored in simpler vs. more complex food webs. Note that mean larval survival is plotted on a natural log scale to reflect the mathematical definition of the adaptive landscape."}
knitr::include_graphics("analyses/MV_landscapes.pdf")
```

@@

<!-- Generality of results -->
The generality of our results likely depends on the relative abundance and functional differences between consumers in a community. For example, if consumers do not differ from each other, then we do not expect changes in food-web complexity to modify selective constraints. Also, many consumers may be at too low of abundances to impose selection on their resources. Rank abundance curves [@Preston1948] and the disproportionate number of weak interactions in diverse communities [@Paine1992] support this notion. When consumers are abundant though, the effect of food-web complexity will depend on whether different species impose conflicting selection pressures or select for distinct traits. For example, parisitoids and birds impose conflicting selection pressures on the size of galls induced by the fly *Eurosta solidaginis* [@Weis1985; @Abrahamson1997; @Start2016]. Recent studies in this system have shown that decreases in the relative abundance of birds, due to either small patch sizes [@Start2016] or proximity to urban areas [@Start2018] causes a shift from neutral to directional selection on gall size. On the other hand, different consumers may impose selection on different traits, favoring trait integration and increasing genetic constraints. Examples of this include strong genetic covariances in plant resistance to different insect herbivores [@Maddox1990; @Wise2007; @Wise2013], although there are also examples where these covariances are weak [@Roche1997; @Barbour2015], or vary from year-to-year [@Johnson2007]. We suggest that gaining predictive insight to the evolutionary consequences of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions. <!--We suggest that gaining predictive insight to the effects of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions.-->

>>>>

```{r setup ms, include=FALSE}
## Reproduce analyses ----
knitr::knit("analyses/reproduce_analyses.Rmd")

@@

We conducted our study within a four-year old common garden experiment of coastal willow (*Salix hookeriana*) located at Humboldt Bay National Wildlife Refuge (HBNWR) (40 40'53"N, 124 12'4"W) near Loleta, California, USA. This common garden consists of 26 different willow genotypes that were collected from a single population of willows growing around Humboldt Bay. Stem cuttings of each genotype (25 replicates per genotype) were planted in a completely randomized design in two hectares of a former cattle pasture at HBNWR. Willows at our study site begin flowering in February and reach their peak growth in early August. During this study, willows had reached 5 - 9m in height. Further details on the genotyping and planting of the common garden are available in @Barbour2015.

@@

We collected data on three different traits that we expected to influence larval survival based on our previous work [@Barbour2016] and other work with gall midges [@Weis1983; @Heath2018]. First, we measured gall diameter as the size of each gall chamber to the nearest 0.01 mm at its maximum diameter (perpendicular to the direction of plant tissue growth). Our previous work has shown that larger galls are associated with higher survival [@Barbour2016]. Second, we measured clutch size by counting the number of chambers in each gall [@Weis1983; @Heath2018]. All larvae collected from the same multi-chambered gall were scored with the same clutch size. Third, we measured oviposition preference as the density of larvae observed on a plant in an independent survey. We did this by randomly sampling five branches per tree and counting the number of individual gall chambers (number of larvae). We then converted these counts to a measure of larval density per 100 shoots by counting the number of shoots on the last branch we sampled. All larvae collected from the same plant were scored with the same oviposition preference. Measuring larval densities on plants in the field is a common method for measuring oviposition preference [@Gripenberg2010]; however, caution must be taken in inferring 'preference' as larval densities can be influenced by processes other than preference [@Singer1986]. Fortunately, a couple features of our study system suggest that larval densities may be a good proxy for oviposition preference. For example, since our data comes from a randomized placement of plant genotypes in a common garden, there is no consistent bias in which plant genotypes females are exposed to while searching for oviposition sites. Also, egg predation is a minor source of mortality for galling insects in general [@Hawkins1997]; therefore, we do not expect any prior egg predation to bias our estimates of observed larval densities.

@@

Our analysis consisted of three parts. First, we used generalized linear mixed models (GLMM) to quantify selection surfaces \textemdash linear and nonlinear relationships between absolute fitness ($W$) and phenotypic traits ($z_i$) of individuals \textemdash in each food-web treatment. Second, we translated selection surfaces into the scale of relative fitness ($w$) in order to calculate selection gradients. Third, we used our estimates of selection gradients to characterize the slope and curvature of the adaptive landscape. Note that inferring adaptive landscapes from selection surfaces assumes that trait distributions are multivariate normal [@Lande1983]. To approximate this assumption, we log-transformed clutch size and square-root transformed oviposition preference. We then scaled all phenotypic traits (mean=0 and SD=1) in order to calculate standardized selection gradients that were comparable across traits and with other studies of natural selection.

@@

All analyses and visualizations were conducted in R [@R2018]. Unless otherwise noted, we report mean estimates of trait-fitness relationships and selection gradients with 95% confidence intervals in brackets. Note that for visualizing the adaptive landscape we restrict trait axes to $\pm 1$ SD of the mean trait value. This emphasizes the fact that we can only reliably estimate the shape of the adaptive landscape near the mean phenotype of the population [@Arnold2001]. We also plot mean larval survival on a natural log scale to accurately reflect the mathematical definition of the adaptive landscape [@Arnold2003]. Data and code to reproduce these analyses is publically available at the GitHub repository: https://github.com/mabarbour/complexity_selection.

@@

\indent We found that more gall midge traits experienced directional selection in the simple (3 of 3) vs. complex food-web treatment (1 of 3)(table \ref{Table:Gradients}). For example, we observed directional selection for smaller clutch sizes in the simple food web, but there was no evidence of selection acting on this trait in the complex food web (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This absence of selection appeared to be a result of conflicting selection pressures imposed by each guild of parasitoids. Specifically, when we subset our data to focus on differences between parasitoid guilds, we found that larval parasitoids actually impose directional selection for larger clutch sizes ($\beta_{\text{Clutch}}$=
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$mean,2)` 
[`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.high,2)`]). In the simple food web, we also observed clear evidence of directional selection for midges to avoid ovipositing on plants with high densities of conspecifics (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}B); however, this relationship was weaker in complex food webs (table \ref{Table:Gradients}). This was likely a result of larval parasitoids imposing greater mortality on egg parasitoids at high gall midge densities (see Selection on the extended phenotype of egg parasitoids section), and thus a less than additive effect on gall midges. Chamber diameter experienced positive directional selection in both food-web treatments, but selection was `r round(((filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean - filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) / filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) * 100, 0)`% higher in complex food webs (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}). This was not due to any difference between egg and larval parasitoids (\alpha_{Diam} contrast = 
`r round(filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$conf.high,2)`]), but was due to the lower average survival in complex food webs (contrast `r round(filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$mean*-1, 2)`
[`r round(filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.low*-1, 2)`,
`r round(filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.high*-1, 2)`]).

@@

\begin{table}[h]
\caption{Standardized selection gradients acting on gall midges in complex vs. simple food webs.}
\label{Table:Gradients}
\centering
\begin{tabular}{lccc}
\\ 
\hline
\textbf{Selection gradient} & \textbf{Complex} & \textbf{Simple} & \textbf{Contrast}  \\ 
\hline
$\beta_{\text{Diam}}$ & 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.high,2)`] }& 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.high,2)`] }& 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.high,2)`] }\\

@@

```{r MV_Landscape, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:MV_Landscape}Two dimensional view of adaptive landscapes of gall midge phenotypes in complex vs. simple food webs. Each panel corresponds to a different combination of phenotypic traits: clutch size and chamber diameter (A); clutch size and oviposition preference (B); oviposition preference and chamber diameter (C). Arrows represent mean estimates of directional selection gradients, while contours represent predicted larval survival of the mean phenotype in each food-web treatment. Notice that arrows point more toward a corner of the adaptive landscape for each combination of traits in the simple vs. complex food-web treatment. This indicates that trait integration (covariance) is more strongly favored in simpler vs. more complex food webs. Note that mean larval survival is plotted on a natural log scale to reflect the mathematical definition of the adaptive landscape."}
knitr::include_graphics("analyses/MV_landscapes.pdf")
```

@@

<!-- Generality of results -->
The generality of our results likely depends on the relative abundance and functional differences between consumers in a community. For example, if consumers do not differ from each other, then we do not expect changes in food-web complexity to modify selective constraints. Also, many consumers may be at too low of abundances to impose selection on their resources. Rank abundance curves [@Preston1948] and the disproportionate number of weak interactions in diverse communities [@Paine1992] support this notion. When consumers are abundant though, the effect of food-web complexity will depend on whether different species impose conflicting selection pressures or select for distinct traits. For example, parisitoids and birds impose conflicting selection pressures on the size of galls induced by the fly *Eurosta solidaginis* [@Weis1985; @Abrahamson1997]. Recent studies in this system have shown that decreases in the relative abundance of birds, due to either small patch sizes [@Start2016] or proximity to urban areas [@Start2018] causes a shift from neutral to directional selection on gall size. On the other hand, different consumers may impose selection on different traits, favoring trait integration and increasing genetic constraints. Examples of this include strong genetic covariances in plant resistance to different insect herbivores [@Maddox1990; @Wise2007; @Wise2013], although there are also examples where these covariances are weak [@Roche1997; @Barbour2015], or vary from year-to-year [@Johnson2007]. We suggest that gaining predictive insight to the evolutionary consequences of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions. <!--We suggest that gaining predictive insight to the effects of food-web disassembly requires an understanding of the mechanisms governing the assembly of trophic interactions.-->
